Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-04-26 response_r2.md

Sincerely,  
Reed A. Cartwright, on behalf of all authors

## Reviewer 1 Comments

1. Comment: *It is now very clear and easy to read manuscript and a important edition to the alignment literature since it does well relative to the objective it pursues.  It has moved away from statistical alignment formulation in the first submission toward doing well in agreeing with benchmark alignments.  Some discussion of this would be useful.*

    Thank you for the positive words. We have added additional language to the manuscript about COATi's success with benchmark alignments.

@@

1. Comment: *And how will the authors turn this into a multiple alignment tool?*

    COATi currently has an early-stage algorithm for estimating pairwise alignments. However, it needs more work (and an additional round of funding) before it meets the high-quality standards exhibited by the pairwise aligner. We have included additional details about COATi MSA in the discussion.

@@

2. Comment: *I am not certain why the authors use “phase-3” indels rather than “phase-0”. I find phase-0 more intuitive, and it strikes me that this could be changed easily and would simplify the writing – the authors wouldn’t have to explain that they using phase-3 instead of phase-0!*

    We originally used phase-0 in our terminology. However, in our experience from presenting this research to diverse audiences, phase-3 was easier terminology to explain than phase-0. For example, "a phase-X gap occurs after the X position in a codon" is easier to explain than "a phase-X gap occurs before the X+1 position in a codon". Importantly, by using phase-3, we can succinctly explain the relationship between the frequencies of the gaps as phase-3 > phase-2 > phase-1. We have added additional language to the manuscript to emphasize these points.

3. Comment: *The authors should use “see” rather than c.f. on p. 5, line 82. The abbreviation should be cf., not c.f., and it should be used to mean “compare.” Given the Ranwez et al. references I don’t think the authors of this paper mean “compare.”*

    We have reworked this section and included additional citations to prior work in this area.

### 2024-04-26 alignpair_letter.tex
Add additional citations.

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citep{sequence_alignment_rosenberg_2009}. While sophisticated advancements have been made, the challenge of alignment inference has not been fully solved \citep{art_morrison_2015}.
%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}. 
%
The alignment of protein-coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \citep[e.g.][]{bininda2005transalign,abascal2010translatorx}. While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts, such as frameshifts and early stop codons. While some aligners can utilize codon substitution models, they are often not robust against coding-sequence artifacts.

Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).

@@

In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{Lunter2005-hk}. An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies. While approaches vary, a statistical aligner for a pair of sequences, $X$ and $Y$, typically finds an alignment, $Aln$, that maximizes the joint probability $P(Aln, X, Y)$ or samples alignments from the posterior $P(Aln | X, Y)$. This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{bradley2007transducers}. Pair-HMMs are computational machines with two output tapes. Each tape represents one sequence, and a path through the pair-HMM represents an alignment of the two sequences. Conceptually, pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.

@@

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; cf.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements both the \cite{muse_gaut_1994} codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[cf.][]{ranwez_macse_2011,ranwez_macse_2018}. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figures/fig-fst-coati.pdf}
\caption{The COATi FST is built from simpler FSTs via composition.
(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M. These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape. The weight of each arc is a conditional probability derived from a codon substitution model. See Fig.~\ref{fig:base-calling} for more details about reading this graph.
%
(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the gap-opening parameter and $e$ is the gap-extension parameter.
Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1. This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions.
%
(c) The COATi FST is derived via composition from the codon substitution, indel, error, and ambiguity FSTs.
}
\label{fig:coati-fst}
\end{figure}

>>>>

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citep{sequence_alignment_rosenberg_2009}. While sophisticated advancements have been made, the challenge of alignment inference has not been fully solved \citep{art_morrison_2015}.
The alignment of protein-coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \citep[e.g.][]{bininda2005transalign,abascal2010translatorx}. While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts, such as frameshifts and early stop codons. While some aligners can utilize codon substitution models, they are often not robust against coding-sequence artifacts.

Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence
\citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.

@@

In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{hein2001algorithm,holmes2001evolutionary,Lunter2005-hk,demaio2021cumulative}. An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies. While approaches vary, a statistical aligner for a pair of sequences, $X$ and $Y$, typically finds an alignment, $Aln$, that maximizes the joint probability $P(Aln, X, Y)$ or samples alignments from the posterior $P(Aln | X, Y)$. This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{holmes2001evolutionary,redelings2007incorporating,problems_cartwright_2009}. Pair-HMMs are computational machines with two output tapes. Each tape represents one sequence, and a path through the pair-HMM represents an alignment of the two sequences. Conceptually, pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.

@@

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs (Figs.~\ref{fig:base-calling} and \ref{fig:coati-fst}), each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements both the \cite{muse_gaut_1994} codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons. (See \citealt{hein1994algorithm,arvestad1997aligning,pedersen1998comparison,ranwez_macse_2011,ranwez_macse_2018} for earlier approaches.) This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figures/fig-fst-coati.pdf}
\caption{The COATi FST is built from simpler FSTs via composition.
(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M. These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape. The weight of each arc is a conditional probability derived from a codon substitution model. See Fig.~\ref{fig:base-calling} for more details about reading this graph.
%
(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the gap-opening parameter and $e$ is the gap-extension parameter.
Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1. This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions \citep[Cf.][]{holmes2001evolutionary,demaio2021cumulative}.
%
(c) The COATi FST is derived via composition from the codon substitution, indel, error, and ambiguity FSTs.
}
\label{fig:coati-fst}
\end{figure}
